|

### 2024-06-06 manuscript/paper.Rmd
Add minor changes to manuscript

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
library(Lahman)
```

@@

In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively. We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique. We further set 
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual. To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in (@OUProcess), the closed form solution for the SDE in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.

@@

Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, we analyze the data via a Bayesian analysis.  Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}). To aid in the sampling procedure for the discrete phenotypes, we declare

@@

# Results {#sec:results}
As previously described, four models for the mean process were considered for each trait.  Within each trait, the model with the lowest DIC was chosen for that train.  All subsequent results presented here use the model with the lowest DIC individually by trait.  Table \ref{whichmodel} shows the model that was selected for each trait.  About half of the traits had models with a trend, and most models did not include an OU component.  The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the onyl model with both a trend and an OU component.  NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.  Explain why this might be and what it means.

@@

## Sexual Dimorphism 
Table \ref{prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.  Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.  These results are shown graphically in figure \ref{fig:prob0-plot}.

@@

## Changes over time
Table \ref{auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}.

@@

```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females. \\label{prob0-table}"}
#Which model was used
mod_df <- data.frame(phenotype = names(unlist(model_list)), mod = unlist(model_list))
mod_code <- data.frame(mod = 0:3, mod_name = c("OU-Trend","No OU-Trend",
                                               "OU-No Trend","NO OU-No Trend"))
results_prob_mu_diff_gt_0_table <- results_prob_mu_diff_gt_0 %>%  mutate(prob_mu_diff_gt_0 = round(prob_mu_diff_gt_0,3)) %>% pivot_wider(names_from = "phenotype", values_from = "prob_mu_diff_gt_0")
library(xtable)
xtable(results_prob_mu_diff_gt_0_table)
```

```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0. \\label{auc-table}"}
results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
#library(xtable)
xtable(results_auc_mu_diff_table)
#knitr::kable(results_auc_mu_diff_table, caption = "The mtcars data.")
```

@@

```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
```

>>>>

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
library(Lahman)
library(bookdown)
```

@@

In the above model descriptions, $\mu_{k,ft_i}$ and $\mu_{k,mt_i}$ model the time-$t_i$ specific mean of phenotype $k$ for female and male stickleback fish, respectively. We point out that, for the discrete traits, the means are represented by $\exp\{\mu_{k,ft_i}\}$ and $\exp\{\mu_{k,mt_i}\}$ for ease of use in our modeling technique. We further set 
\begin{align}
\mu_{k,gt_i} = \beta_{0,kg} + \beta_{1,kg}t_i + u_{k,gt_i},
\label{eq:mu}
\end{align}
for $g \in \{f,m\}$ where $\beta_{0,kg}$ and $\beta_{1,kg}$ are regression parameters of phenotype $Y_k$ for each gender, accounting for the possibility of a time-dependent trend in the mean structure, and $u_{k,gt_i}$ is the corresponding residual. To account for potential correlations between the residuals for a given trait $k$ and gender $g$, we fit an Ornstein-Uhlenback (OU) process (@OUProcess). More specifically, define $du_{k,gt} = u_{k,g(t + dt)} - u_{k,gt}$, the change in $u_k,gt$ for a given trait $k$ and gender $g$ over a miniscule time period $dt$. The OU process is defined as 
\begin{align}
du_{k,gt} = -\kappa_k u_{k,gt} dt + \tau_k dW_t,
\label{eq:OU_cont}
\end{align}
where $\kappa_k$ is a parameter associated with the correlation between $u_{k,gt}$ and $u_{k,g(t+dt)}$, $\tau_k$ is the standard deviation of the OU process, and $W_t$ is a standard Brownian motion. As shown in @OUProcess, the closed form solution for the SDE in (\ref{eq:OU_cont}) is
\begin{align}
u_{k,gt_i} \iid \mathcal{N}\left(u_{k,gt_{i-1}}\exp\{-\kappa_k(t_{i} - t_{i-1})\} , \frac{\tau_k^2(1 - \exp\{-2\kappa_k(t_{i} - t_{i-1}))\}}{2\kappa_k}\right)
\label{eq:OU_sol}
\end{align}
for $i = 2,\cdots,T$. In a traditional OU process, the initial value $u_{k,gt_1}$ is assumed to be a (potentially unknown) constant, and we discuss the procedure for estimating this value below.

@@

Because we are fitting a dataset with a stochastic structure on the means of the phenotypes, and we are interested in the distribution of the overall structure of these means, we analyze the data via a Bayesian analysis.  Bayesian data analysis is also more naturally used when we have to impute data (\textbf{CITATION}). To aid in the sampling procedure for the discrete phenotypes, we declare

@@

# Results {#sec:results}
As previously described, four models for the mean process were considered for each trait.  Within each trait, the model with the lowest DIC was chosen for that train.  All subsequent results presented here use the model with the lowest DIC individually by trait.  Table \ref{whichmodel} shows the model that was selected for each trait.  About half of the traits had models with a trend, and most models did not include an OU component.  The traits that did include an OU component were caudal vertebra (mcv), pelvic spine lentgh (lpt), abdominal vertebra (mav), and standard length (stl) with caudal vertebra (mcv) being the only model with both a trend and an OU component.  This implies that not only does caudal vertebra size have a linear trend across gender, but the average size at one time period is highly correlated with the average size at the next time period.<!-- NOTE FOR MATT: You should mention something about MCV being the only one with a trend and an OU.  Explain why this might be and what it means.  -->

@@

## Sexual Dimorphism 
Table \ref{tab:prob0-table} shows the posterior probability of the differences in the means between male and female specimens at each time point.  Probabilities near 0.5 indicate very little difference in the means whereas probabilities far from 0.5 indicate sexual dimorphism with values of 0 and 1 indicating larger mean values for females and males, respectively.  These results are shown graphically in figure \ref{fig:prob0-plot}.

@@

## Changes over time
Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}.

@@

```{r prob0-table, echo = FALSE, tab.cap= "Posterior probability that the mean for males is larger than the mean for females"}
#Which model was used
mod_df <- data.frame(phenotype = names(unlist(model_list)), mod = unlist(model_list))
mod_code <- data.frame(mod = 0:3, mod_name = c("OU-Trend","No OU-Trend",
                                               "OU-No Trend","NO OU-No Trend"))
results_prob_mu_diff_gt_0_table <- results_prob_mu_diff_gt_0 %>%  mutate(prob_mu_diff_gt_0 = round(prob_mu_diff_gt_0,3)) %>% pivot_wider(names_from = "phenotype", values_from = "prob_mu_diff_gt_0")
library(xtable)
xtable(results_prob_mu_diff_gt_0_table)
```

```{r auc-table, echo = FALSE, tab.cap= "AUC comparing time x versus time 0."}
results_auc_mu_diff_table <- results_auc_mu_diff %>%  mutate(auc_mu_diff = round(auc_mu_diff,3)) %>% pivot_wider(names_from = "phenotype", values_from = "auc_mu_diff")
#library(xtable)
xtable(results_auc_mu_diff_table)
#knitr::kable(results_auc_mu_diff_table, caption = "The mtcars data.")
```

@@

```{r prob0-plot, echo = FALSE, fig.cap = "Posterior probabilities that mean for males is greater than mean for females"}
ggplot(aes(x = time, y = prob_mu_diff_gt_0), data = results_prob_mu_diff_gt_0) + facet_wrap(~phenotype) + geom_path() + #geom_hline(yintercept = 0, lty = 3) + ylim(0,1) + ylab("Posterior Probability") Why yintercept in 0 and not 0.5?
geom_hline(yintercept = 0.5, lty = 3) + ylim(0,1) + ylab("Posterior Probability")
```
